Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-08-31 Master Thesis/introduction.tex
Chapter restructuring

In the first chapter, we want to recap some results from valuation theory. In the second part, we will introduce some lemmata in order to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.

In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields. Using the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding discrete valuations. We will proceed to show that $\sigma$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.

In the third chapter, we want to use the correspondences established in the second chapter to prove Theorem \ref{thm:main-result}. Before we are able to do that, we need to discuss free-by-finite groups. Using the existence of a so-called admissible set, we can construct a group isomorphism $K'^\times\to K^\times$. Finally, we will show that this group isomorphism can be extended to a field isomorphism $K'\to K$. By passing to the limits, we obtain a field isomorphism $\varphi: \overline{K'}\to\overline{K}$ which induces $\sigma$.

>>>>

In the first chapter, we want to recap some results from valuation theory. Then we will introduce some lemmata in order to be able to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.

In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields. Using the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also see that $\sigma$ induces a field isomorphism between the residue fields of corresponding discrete valuations.

In the third chapter, we will establish a field isomorphism between the constant fields of $K$ and $K'$ respectively. We will proceed to show that $\sigma$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.

In the fourth chapter, we want to use the correspondences established in the third chapter to prove Theorem \ref{thm:main-result}. Before we are able to do that, we need to discuss free-by-finite groups. Using the existence of a so-called admissible set, we can construct a group isomorphism $K'^\times\to K^\times$. Finally, we will show that this group isomorphism can be extended to a field isomorphism $K'\to K$. By passing to the limits, we obtain a field isomorphism $\varphi: \overline{K'}\to\overline{K}$ which induces $\sigma$.

### 2017-08-31 Master Thesis/synopsis.tex
Chapter restructuring

\chapter{Partial Correspondences}

@@

\section{Local Correspondence}

In this section we want to see that $\sigma$ induces bijections on defectless valuations. A direct consequence of Proposition \ref{2.6} is the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$:

@@

Let $n\in\mathbb{N}$. In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$. By Proposition \ref{prop:kummer-theory}, we have isomorphisms:
\[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]
In this section, our goal is to establish an isomorphism between the corresponding cohomology groups.

\begin{definition}
Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e. the constant fields of $K$ and $K'$, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closures of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.

@@

By Proposition \ref{3.2} $\sigma$ induces a bijection $\sigma: X^{\phantom{'}}_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.
\end{definition}

@@

We start by showing that $\sigma$ induces an isomorphism between the absolute Galois groups of the constant fields $G_k$ and $G_{k'}$. We need the following Lemma:

@@

\begin{lemma}\label{4.2}
We have $(K\overline{k})'=K'\overline{k'}$, i.e. $\sigma(G_{K\overline{k}})=G_{K'\overline{k'}}$. $\sigma$ induces an isomorphism $G_k\to G_{k'}$ such that the following diagram commutes:
\[ \begin{tikzcd}
G_K \ar[r, "\sigma"] \ar[d] & G_{K'}\ar[d]\\
G_k \ar[r, "\sigma"'] & G_{k'}
\end{tikzcd} \]
Furthermore, $\sigma:G_k\to G_{k'}$ is induced by a unique isomorphism $\varphi:\overline{k'}\to\overline{k}$ which restricts to an isomorphism $\varphi: k'\to k$.
\end{lemma}

\begin{proof}
Let $L/K$ be a finite subextension of $\overline{K}/K$. Let $\ell$ and $\ell'$ be the constant fields of $L$ and $L'$ respectively. Applying Lemma \ref{4.2-pre} to $K$ and $L$ gives:
\[ [\ell:k] = [\ell':k'] \]
Therefore, $L/K$ is a constant field extension if and only if $L'/K'$ is a constant field extension. Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$. This implies that $\sigma$ induces an isomorphism:
\[G_k \cong \operatorname{Gal}(K\overline{k}/K) \to \operatorname{Gal}(K'\overline{k'}/K')\cong G_{k'}\] 
Everything else follows from Neukirch-Uchida \ref{thm:neukirch-uchida}.
\end{proof}

\begin{remark}
Let $\overline{K''} = \overline{K'}\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
\[ \begin{tikzcd}
G_K \ar[r, "\sigma"] & G_{K'} \ar[r, "\phi", "\cong"'] & \operatorname{Gal}(\overline{K''}/K')
\end{tikzcd} \]
where $\phi$ is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and that $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities. Now the following diagram commutes:
\[ \begin{tikzcd}
G_K \ar[d]\ar[r, "\sigma"] & G_{K'} \ar[ld]\\
G_{k} &
\end{tikzcd} \]
\end{remark}

\paragraph{} Let $n\in\mathbb{N}$. The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]

@@

\chapter{Theorem}

Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us construct a group isomorphism $\varphi: K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$ we constructed in the last chapter.

\section{Free-By-Finite Groups}

@@

\begin{proof}
Let $P_1\in X_0-S$. We set:
\[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
Since $r_{P_1}|_{k^\times}:k^\times \to k(P_1)^\times$ is injective, we have $C_1\cap k^\times = 1$ and the following map is injective:
\[ C_1 \to\prod_{P\in S}\mathbb{Z},\ x \mapsto (v_P(x))_{P\in S} \]
Therefore, $C_1$ is a free abelian group of finite rank. If $C_1 = 1$, then we set $T = \{P_1\}$ and we are done. Otherwise, let $x\in C_1,\ x\neq 1$. Then there exists an $a\in k^\times$ with the following properties:
\begin{enumerate}[(i)]
\item There exists a point $P_2\in X_0 - (S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
\item $a$ is not a root of unity.
\end{enumerate}
Clearly, (i) and (ii) hold for almost all $a\in k^\times$. We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$. $C_2$ is free abelian of finite rank as well. By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$. By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in X_0-S$ such that the following map is injective:
\[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
Finally for $T=\{P_1,\ldots,P_r\}\subset X_0-S$ the map $r_{S,T}$ is injective. For the second assertion note the following isomorphism:
\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
Since $K^S/k^\times$ is generated by the uniformizers of the corresponding valuations in $S$, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
\end{proof}

@@

\section{Proof of Pop's Theorem}

>>>>

\chapter{Local Correspondence}

@@

In this chapter we want to see that $\sigma$ induces bijections on defectless valuations. A direct consequence of Proposition \ref{2.6} is the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$:

@@

\chapter{Global Correspondence}

Let $k$ and $k'$ be the algebraic closures of $\mathbb{Q}$ in $K$ and $K'$ respectively, i.e. the constant fields of $K$ and $K'$, and let $\overline{k}$ and $\overline{k'}$ be the algebraic closures of $k$ and $k'$ in $\overline{K}$ and $\overline{K'}$ respectively.

@@

By Proposition \ref{3.2} $\sigma$ induces a bijection $\sigma: X^{\phantom{'}}_0\to X'_0,\ P\mapsto P'$ and for any $P\in X_0$ an isomorphism $\varphi_P: k(P')\to k(P)$.

\section{Isomorphism of Constant Fields}

In this section we shall establish an isomorphism between the absolute Galois groups of the constant fields $k$ and $k'$. In particular, since both $k$ and $k'$ are number fields, the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida} will yield an isomorphism between $k$ and $k'$.

@@

@@

\begin{lemma}\label{4.2}
We have $(K\overline{k})'=K'\overline{k'}$, i.e. $\sigma(G_{K\overline{k}})=G_{K'\overline{k'}}$. $\sigma$ induces an isomorphism $\sigma: G_k\to G_{k'}$ such that the following diagram commutes:
\[ \begin{tikzcd}
G_K \ar[r, "\sigma"] \ar[d] & G_{K'}\ar[d]\\
G_k \ar[r, "\sigma"'] & G_{k'}
\end{tikzcd} \]
Furthermore, $\sigma:G_k\to G_{k'}$ is induced by a unique isomorphism $\varphi:\overline{k'}\to\overline{k}$ which restricts to an isomorphism $\varphi: k'\to k$.
\end{lemma}

\begin{proof}
Let $L/K$ be a finite subextension of $\overline{K}/K$. Let $\ell$ and $\ell'$ be the constant fields of $L$ and $L'$ respectively. Applying Lemma \ref{4.2-pre} to $K$ and $L$ gives:
\[ [\ell:k] = [\ell':k'] \]
Therefore, $L/K$ is a constant field extension if and only if $L'/K'$ is a constant field extension. Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$. This implies that $\sigma$ induces an isomorphism:
\[G_k \cong \operatorname{Gal}(K\overline{k}/K) \to \operatorname{Gal}(K'\overline{k'}/K')\cong G_{k'}\] 
Everything else follows from the theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}.
\end{proof}

Let $n\in\mathbb{N}$. In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$. By Proposition \ref{prop:kummer-theory}, we have isomorphisms:
\[ \mathrm{H}^1(K',\mu_n)\cong K'^\times/n,\qquad \mathrm{H}^1(K,\mu_n)\cong K^\times/n \]
Our goal is to establish an isomorphism between these cohomology groups.

Let $\overline{K''} = \overline{K'}\otimes_{\overline{k'}}\overline{k}$ via the isomorphism $\varphi: \overline{k'}\to \overline{k}$. If we replace $\sigma: G_K\to G_{K'}$ with the composition
\[ \begin{tikzcd}
G_K \ar[r, "\sigma"] & G_{K'} \ar[r, "\phi", "\cong"'] & \operatorname{Gal}(\overline{K''}/K')
\end{tikzcd} \]
where $\phi$ is induced by the isomorphism $\overline{K''}\to\overline{K'},\ x\otimes y\mapsto \varphi^{-1}(y)\cdot x$, we may assume that $k=k'$, $\overline{k} = \overline{k'}$ and that $\varphi:\overline{k}\to\overline{k}$, $\sigma: G_k\to G_k$ are identities. Now the following diagram commutes:
\[ \begin{tikzcd}
G_K \ar[d]\ar[r, "\sigma"] & G_{K'} \ar[ld]\\
G_{k} &
\end{tikzcd} \]

The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\zeta = g\zeta$ for all $\zeta\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]

@@

\chapter{Proof of Pop's Theorem}

Before we start proving Pop's Theorem \ref{thm:main-result}, we need some tools that will help us construct a group isomorphism $\varphi: K'^\times\to K^\times$ from the modulo $n$ isomorphisms $\varphi_n: K'^\times/n\to K^\times/n$ which we constructed in the previous chapter.

\section{Free-by-Finite Groups}

@@

\begin{proof}
Let $P_1\in X_0-S$. We set:
\[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
Since $r_{P_1}|_{k^\times}:k^\times \to k(P_1)^\times$ is injective, we have $C_1\cap k^\times = 1$ and the following map is injective:
\[ C_1 \to\prod_{P\in S}\mathbb{Z},\ x \mapsto (v_P(x))_{P\in S} \]
Therefore, $C_1$ is a free abelian group of finite rank. If $C_1 = 1$, then we set $T = \{P_1\}$ and we are done. Otherwise, let $x\in C_1,\ x\neq 1$. Then there exists an $a\in k^\times$ with the following properties:
\begin{enumerate}[(i)]
\item There exists a point $P_2\in X_0 - (S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
\item $a$ is not a root of unity.
\end{enumerate}
Clearly, (i) and (ii) hold for almost all $a\in k^\times$. We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$. $C_2$ is free abelian of finite rank as well. By requirement, $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$. By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in X_0-S$ such that the following map is injective:
\[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
Finally for $T=\{P_1,\ldots,P_r\}\subset X_0-S$ the map $r_{S,T}$ is injective. For the second assertion note the following isomorphism:
\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
Since $K^S/k^\times$ is generated by the uniformizers of the corresponding valuations in $S$, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
\end{proof}

@@

\section{Proof}
